# Proper Loans Explanatory Data Analysis
### Wafic Fahme

-------------------------------------------------------------------------------

# Introduction

-------------------------------------------------------------------------------

Why Prosper:

Prosper is a leading P2P in the lending industry and it attracted me most \
because I work in the financial sector and companies like Lending Club and \
Prosper are interesting to us as we read a lot about them in the media.

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
library(ggplot2)
library(dplyr)
library(gridExtra)
library(GGally)

```

```{r echo=FALSE, Load_the_Data}
loan_data <- read.csv("/Users/wafic/Github/R/Final_project_R/prosperLoanData.csv"
                      , na.strings=c("","NA"), as.is=TRUE)
```

-------------------------------------------------------------------------------

# Univariate Plots

-------------------------------------------------------------------------------

```{r echo=FALSE}
dim(loan_data)
```

In the prosper data set provided we have 81 variables and 113,937 \
observation.


```{r echo=FALSE, warning=FALSE, message = FALSE}
ggplot(aes(x = BorrowerAPR), data = loan_data)+
  geom_histogram()
```

We can see that the APR every borrower of the 114,000 loans we have payed.

```{r echo=FALSE, warning=FALSE, message = FALSE}
ggplot(aes(x = BorrowerRate), data = loan_data)+
  geom_histogram()
```

It looks very similar to APR in shape and distribution

```{r echo=FALSE}
ggplot(aes(x = LoanStatus), data = loan_data)+
  geom_bar()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

Almost half of the loans are still in process, and around 4,000 of the loans \ 
have been completed, but we can see almost 11,000 charged off their loans and  \
almost 5,000 borrowers defaulted on their loans.

```{r echo=FALSE}
loan_data$LoanStatus <- factor(loan_data$LoanStatus)
table(loan_data$LoanStatus)
```

Number of loans in the every status, we have around 12,000 loans of the \
114,000 observations charged off

```{r echo=FALSE, warning=FALSE, message = FALSE}
ggplot(aes(x = Term), data = loan_data)+
  geom_bar()
```

We can seee that the majority of loans (around 80,000) fall in the middle \
which gets us to assume that usually loan amounts has to be small to be paid \ 
in such a short term

Loan Terms
```{r echo=FALSE}
table(loan_data$Term)
```
We have 3 groups of borrowers here, with around 88,000 in the 36 month period \
(3yrs) and period and the second popular is 60 month period (5yrs) with 25,000 \
borrowers and least is the 12 months which I assume is for borrowers of small \
amounts

```{r echo=FALSE, warning=FALSE}
ggplot(aes(x = LoanOriginalAmount), data = loan_data)+
  geom_histogram(binwidth = 500)
```

We can see spikes at some specific loan amount like 10,000, 20,000 and 25,000 \
Lets break it see the amount more clear

Loan Original Amount
```{r echo=FALSE, warning=FALSE}
summary(loan_data$LoanOriginalAmount)
```

We can see that the median of a loan at prosper is 6,500 which confirms our \
earlier expectations of why the loan terms are mostly 36 month period. \
The number of borrowers in the Loan Original Amount is right skewed, although \ 
we can see spikes at amounts: \
- $4,000 (15,000 borrowers)
- $5,000 (7,500 borrowers)
- $10,000 (13,000 borrowers)
- $15,000 (15,000 borrowers)
- $20,000 (3,000 borrowers)
- $30,000 (3,000 borrowers)
This makes sense because people will take a loan for 15,000 if they want around \
14,000 and when the figures become bigger we see less and less variations in \ 
the amounts borrowed unlike the smaller amounts when 1,000 makes a difference \
for someome wanting to borrow 2,000.

```{r echo=FALSE, Univariate_Plots_5}
loan_data$ListingCreationYear <- 
  format(as.Date(loan_data$ListingCreationDate), "%Y")

ggplot(aes(x = ListingCreationYear), data = loan_data)+
  geom_bar()
```

If we look at years 2008-2009, we will notice that in 2008 the growth of the \
number of borrowers stall and then dramatically dropped due to the housing \
crisis, but then it started growing in a fast pace. In 4 years the number of \
loans went from 5,000 in 2010 to 35,000 loans in 2013.


```{r echo=FALSE, Univariate_Plots}
ggplot(aes(x = IncomeRange), data = loan_data)+
  geom_bar()
```

Around 60,000 borrowers salary ranges between $25,000 and $75,000 but the \
surprise is in the numbers of borrowers whom salary is $100,000 and above \
whom are around 17,000 borrowers, I wonder how much they borrowed especially \
that the amount borrowed median is $6,500, are they the borrowers of the \
$30,000 we saw earlier.

```{r echo=FALSE}
ggplot(data = loan_data, aes(x = BorrowerState))+
  geom_bar()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

We can see that the highest state in terms of prosper borowers is CA which \
is expected because such P2P startups are still in the early adopters stage \
and usually states like California tends to have have cultures that are more \
adoptive of online and new tech ideas.


```{r echo=FALSE}
ggplot(aes(x = EmploymentStatus), data = loan_data)+
  geom_bar()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

Almost above 90,000 of the borrowers are either employed and very little \
are un-employed or retired, I wonder how much they have borrowed for how long \

```{r echo=FALSE, warning=FALSE, message = FALSE}
ggplot(aes(x = MonthlyLoanPayment), data = loan_data)+
  geom_histogram()
```

We can see that the majority of borrowers have a monthly loan payment \
less than $300, even 30,000 of the borrowers pay as low as arounf $150 per \
month


```{r echo=FALSE, warning=FALSE}
loan_data$CurrentDelinquencies <- factor(loan_data$CurrentDelinquencies,
                                             levels = c('0','1', '2',
                                                        '3', '4', '5',
                                                        '6', '7', '8',
                                                        '9', '10'))

ggplot(data = loan_data, aes(x = CurrentDelinquencies))+
  geom_bar()
```

After factoring and leveling the data, we have almost 90,000 borrowes of our \
113,000 has no deliquencies whatsoever with around 12,000 whom have 1 account \
delinquent.

-------------------------------------------------------------------------------

# Univariate Analysis

-------------------------------------------------------------------------------

### What is the structure of your dataset?

The data set has 81 variables, but I have chosen these to work with for my EDA.

* ListingCreationDate:	The date the listing was created.

* Term:	The length of the loan expressed in months.

* LoanStatus:	The current status of the loan: Cancelled,  Chargedoff, Completed,
Current, Defaulted, FinalPaymentInProgress, PastDue. The PastDue status will be 
accompanied by a delinquency bucket.

* BorrowerAPR:	The Borrower's Annual Percentage Rate (APR) for the loan.

* BorrowerRate:	The Borrower's interest rate for this loan. 

* ProsperRating (Alpha):	The Prosper Rating assigned at the time the listing 
was created between AA - HR.  Applicable for loans originated after July 2009.

* ProsperScore:	A custom risk score built using historical Prosper data. 
The score ranges from 1-10, with 10 being the best, or lowest risk score.  
Applicable for loans originated after July 2009.

* CreditScoreRangeUpper:	The upper value representing the range of the 
borrower's credit score as provided by a consumer credit rating agency. 

* AmountDelinquent:	Dollars delinquent at the time the credit profile was pulled.

* BankcardUtilization:	The percentage of available revolving credit that is 
utilized at the time the credit profile was pulled.

* DebtToIncomeRatio:	The debt to income ratio of the borrower at the time the 
credit profile was pulled. This value is Null if the debt to income ratio is 
not available. This value is capped at 10.01 (any debt to income ratio larger 
than 1000% will be returned as 1001%).

* IncomeRange:	The income range of the borrower at the time the listing was 
created.

* StatedMonthlyIncome:	The monthly income the borrower stated at the time the 
listing was created.

* LoanOriginalAmount:	The origination amount of the loan.

* Investors:	The number of investors that funded the loan.

* MonthlyLoanPayment	The scheduled monthly loan payment.

* BorrowerState:	The two letter abbreviation of the state of the address of 
the borrower at the time the Listing was created.

Create a new dataframe with the above variables
```{r echo=FALSE, create_new_data_frame}
prosper_loans <- select(loan_data, c(ListingCreationDate, Term, LoanStatus, 
                                     BorrowerAPR, 
                                     BorrowerRate, ListingKey, MonthlyLoanPayment,
                                     ProsperRating..Alpha.,ProsperScore, 
                                     BorrowerState,
                                     CreditScoreRangeUpper, AmountDelinquent,
                                     BankcardUtilization,DebtToIncomeRatio, 
                                     IncomeRange, StatedMonthlyIncome,
                                     LoanOriginalAmount, Investors))
```


```{r echo=FALSE}
prosper_loans$LoanStatus <- factor(prosper_loans$LoanStatus,
                                   levels = c('Current', 'Completed',
                                              'Chargedoff','Cancelled',
                                              'Defaulted', 'FinalPaymentInProgress',
                                              'Past Due (1-15 days)',
                                              'Past Due (16-30 days)',
                                              'Past Due (31-60 days)',
                                              'Past Due (61-90 days)',
                                              'Past Due (91-120 days)',
                                              'Past Due (>120 days)'))

prosper_loans$Term <- factor(prosper_loans$Term, levels = c('12', '36', '60'))

prosper_loans$ProsperRating..Alpha. <- factor(prosper_loans$ProsperRating..Alpha.,
                                              levels = c("AA", "A", "B", "C", "D",
                                                         "E", "HR"))

prosper_loans$IncomeRange <- factor(prosper_loans$IncomeRange, levels = 
                                        c("Not employed", "$0", "$1-24,999", 
                                          "$25,000-49,999",
                                          "$50,000-74,999", "$75,000-99,999",
                                          "$100,000+"
                                          ))

prosper_loans$BorrowersGroups <- ifelse(prosper_loans$LoanStatus == "Current" |
                                        prosper_loans$LoanStatus == "Completed" |
                                        prosper_loans$LoanStatus == "Past Due (1-15 days)" |
                                        prosper_loans$LoanStatus == "Past Due (16-30 days)"|
                                        prosper_loans$LoanStatus ==  "FinalPaymentInProgress",
                                        "Non-Delinquent", "Delinquent")

prosper_loans$BorrowersGroups <-factor(prosper_loans$BorrowersGroups, levels = 
                                            c("Non-Delinquent", "Delinquent"))

prosper_loans$ListingCreationYear <- 
  format(as.Date(prosper_loans$ListingCreationDate), "%Y")

prosper_loans <- select(prosper_loans, -c(ListingCreationDate))

names(prosper_loans)
```


```{r echo=FALSE}
levels(prosper_loans$LoanStatus)
```

I will group Current, Completed, Past Due (1-15 days), Past Due (16-30 days) \
and "FinalPaymentInProgress" \
as Non-Delinquent because a borrower who is only few days late on his or her \
payment might have reasons that are quite random an not covered in the data \
we have. Then I will group all the 8 levels into delinquent borrowers \
because for example a borrower whom is 90 days past due is probably gonna \ 
cancel or default on his loan and I believe they might also be delinquent \
elsewhere as we will try to investigate via correlating the status with the \ 
number of delinquencies a borrower has.

```{r echo=FALSE}
ggplot(aes(x = BorrowersGroups), data = prosper_loans)+
  geom_bar()
```


```{r echo=FALSE}
str(prosper_loans)
```

### What is/are the main feature(s) of interest in your dataset?
I was suprised to see a varying APR levels for every borrower and I was\
curious to undertsand why the difference. \
I want investigate why APR levels changes from loan to another, what are the \
factors that get me as a borrower to have a lower APR at prosper. I hope the \
variables will help me understand the reason beind the difference in rates. \

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?
We will look into the financial status of the borrower thus the borrower \
EmploymentStatus, CreditScore, AmountDelinquent, BankcardUtilization. Also, \ 
we will look at his status with Prosper.com, what his ProsperScore is, \
Term and ListingCreationYear.

### Did you create any new variables from existing variables in the dataset?

I created two variables:

* ListingCreationYear: This is the year the loan was created, to consolidate \
loans as they are currently listed in full date.
* BorrowersGroups: This is the variable meantioned earlier that groups \
borrowes into two groups, where Current, Completed, Past Due (1-15 days) and \ 
Past Due (16-30 days) are Non-Delinquent and the rest are delinquent borrowers.


### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

Knowing that every observation is a loan record, I run the below code to \
notice that there are some duplicated that needs to be removed.

```{r}
length(prosper_loans$ListingKey)
length(unique(prosper_loans$ListingKey))

# Removed all the duplicate loans listings
prosper_loans <- prosper_loans[!duplicated(prosper_loans$ListingKey), ]
length(prosper_loans$ListingKey)
```

```{r echo=FALSE}
# Remove variables I will not use
prosper_loans <- select(prosper_loans, -c(ListingKey))
prosper_loans <- select(prosper_loans, -c(LoanStatus))
dim(prosper_loans)
```


Ended up with 17 variables and 113,066 observations to be investigated 

-------------------------------------------------------------------------------

# Bivariate Plots Section

-------------------------------------------------------------------------------

```{r echo=FALSE, warning=FALSE, message = FALSE}
# sample 10,000 loans from the data set
#prosper_loans_subset <- select(prosper_loans, c(Term, BorrowerAPR, 
 #                                    BorrowerRate,
  #                                   ProsperRating..Alpha.,ProsperScore,
   #                                  CreditScoreRangeUpper, AmountDelinquent,
    #                                 BankcardUtilization,DebtToIncomeRatio, 
     #                                StatedMonthlyIncome, LoanOriginalAmount))
#set.seed(339487731)
#prosper_loans_samp <- prosper_loans_subset[sample(1:length(prosper_loans$BorrowerAPR), 10000), ]
#ggpairs(prosper_loans_samp,
#  lower = list(continuous = wrap("points", shape = I('.'))),
#  upper = list(combo = wrap("box", outlier.shape = I('.'))))
```


```{r echo=FALSE,  warning=FALSE, message = FALSE}
ggcorr(prosper_loans[, 0:17], label = TRUE)
```

Looking at the correlation coofecient in the relationship between every \
varaible we are investigating in Prosper, we can see that some variables like \ 
Borrower APR and Borrower Rate are highly correlated but if we think of it, we \ 
know for a fact that these are identical variables and the correlation is due \
to the fact that they are identical. We can also see a high correlation \
between loan amount and the monthly payment which makes sence as the payments \
increase with the size of the loan. We can also see a rather strong negative \
correlation between the Borrower AP and the Prosper Score which can be a start \
to our hypothesis regarding that Prosper gives high APR scores to risky \
customers.


```{r echo=FALSE, warning=FALSE}
ggplot(aes(x = BorrowersGroups, y = BorrowerAPR), data = prosper_loans)+
  geom_boxplot()
```

In trying to understand the variation of the APR between borrowers, I \
started with the delinquent vs non-delinquent and its obvious that the \
delinquent group pages on average a much higher rate of 25% vs 20% for the \ 
non-delinquent group and since APR rates are imposed at the beginning of the \
loan, this gets us to guess the prosper might have believed that these \
borrowers are riskier.

```{r echo=FALSE, warning=FALSE}
ggplot(aes(x = BorrowerRate, y = BorrowerAPR), data = prosper_loans)+
  geom_point(alpha=1/10)
```

This make sense because we know that borrower rate and APR are complimentary \
variables, but what we dont see is a perfect correlation between the 2 \
variables, for example a borrower can pay a rate of 20% and an APR between \
20% and 30% which is quite a difference. We can also spot some outlier paid \
borrower rate less than 10% but an APR of almost 22%.

```{r echo=FALSE, warning=FALSE}
ggplot(aes(x = ListingCreationYear, y = BorrowerAPR), data = prosper_loans)+
  geom_boxplot()
```

Looking on how the APR changed over time, we see that the mean was high \
then dropped then reached its peak in 2011 then dropped to its lowest rate \ 
ever in 2014. I have a hypothesis for it and it has to do that during 2008 and \ 
2011 where the increase in APR is visible, due to the housing crisis the \
borrowers where riskier, for that Prosper had to increase its APR to protect \
its investors and maybe even attract investors as these where tough times.

```{r echo=FALSE, warning=FALSE}
ggplot(aes(x = LoanOriginalAmount, y = BorrowerAPR), data = prosper_loans)+
  geom_point(alpha = 1/5)
```

I expected this plot to have a linear relation, but what I can see is that \
a borrower can of an amount can have a huge variation in his/her APR.\
We can vertical lines due to the nature of the loan amounts as they are usually\
complete numbers like 20,000 or 10,000 etc but the numbers between them I \
guess is due to the fact that the loan was not fully funded by investors.\
For example, if you want to borrow $25,000 you might pay as low as 5% and as \
high as 35% for the same loan amount

```{r echo=FALSE, warning=FALSE}
ggplot(aes(x = Term, y = BorrowerAPR), data = prosper_loans)+
  geom_boxplot()
```

From this plot we can see that the 3 terms have very similar median, and \
even in the same term we can see that the APR can range from 15% to 30% on the \
36 month term (the most popular term of 80,000 borrowers)

```{r echo=FALSE, warning=FALSE}
ggplot(aes(x = ProsperRating..Alpha., y = BorrowerAPR), data = prosper_loans)+
  geom_boxplot()
```

The Prosper rating is the rating given day one for the loan, which I am not \
sure if it affects the APR or vise versa but what I can see it a strong \
relationship between this categorical variable and the APR, because as the \ 
loan becomes riskier(AA least HR highest) the average APR rate jumps from under \
10% to as high as 35% although this graph is full of outliers. \ 
If we look at the lowest APR rate for HR (riskiest borrowers) we can see an APR \
of 20% which is very similar to the median of the much less risky "B" rated \
borrowers.

```{r echo=FALSE, warning=FALSE}
ggplot(aes(x = IncomeRange, y = BorrowerAPR), data = prosper_loans)+
  geom_boxplot()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

One of the most obvious variables will be a borrower salary, and it is as \
anticipated, borrowers of low salaries have a median of 25%, while we can see \
that the higher a person salary the lower he get in APR on average, but I am \
again suprised with this outlier of 100,000+ salary who had an APR of 40%.


```{r echo=FALSE, warning=FALSE}
ggplot(aes(x = StatedMonthlyIncome, y = BorrowerAPR), data = prosper_loans)+
  geom_point()
```

> We can see this borrower here, he has a salary above 1,500,000 per month and
unless Bill Gates is asking for a prosper loan, I believe this is a user entry 
error or maybe a data wrangling issue. For that I will mutate salaries above 
250,000 and replot the above.

Changing salaries above 200,000 by dividing them by 100
```{r echo=FALSE}
prosper_loans$StatedMonthlyIncome <- ifelse(prosper_loans$StatedMonthlyIncome > 200000,
                                            prosper_loans$StatedMonthlyIncome/100,
                                            prosper_loans$StatedMonthlyIncome)
summary(prosper_loans$StatedMonthlyIncome)

```

Plotting the monthly salaries after doing the above treatment that removed
most of the outliers

```{r echo=FALSE, warning=FALSE}
ggplot(aes(x = StatedMonthlyIncome, y = BorrowerAPR), data = prosper_loans)+
  geom_jitter(alpha = 2/5, shape = 21, fill = I("#FFA500"))+
  scale_x_log10()
```

Since the data is concentrated to the left, I experimented it with log10 and \
now the data plot is more clear, we can't still see the relationship between \
the monthly salary and the APR levels, but its now clear that the majority of \
borrowers are in the same financial situation with average salary of 5,000. \
We can see a 2 vertical lined on zero and near zero, and I guess this is either \
missing data or non-employed borrowers. We can see that themost popular APR \
is of 35% as its spread along the monthly salaries from 100 to 12,000.

```{r echo=FALSE, warning=FALSE, message = FALSE}
ggplot(aes(x = DebtToIncomeRatio, y = BorrowerAPR), data = prosper_loans)+
  geom_point()
```

Expected a linear relation here as well, but I can conclude that most borrowers \
are under 1% DTI and some are in the 10% DTI but lets replot it with log10 to \
clear the congestion of points on the left

```{r echo=FALSE, warning=FALSE, message = FALSE}
ggplot(aes(x = DebtToIncomeRatio, y = BorrowerAPR), data = prosper_loans)+
  geom_jitter(alpha = 2/5, shape = 21, fill = I("#FFA500"))+
  scale_x_log10()
```

Similar to amount borrowed we can see that DTI is plotting vertical lines \
that represent the ratio, which is highly concentrated between 0.2 and 0.3 \
but here we can't say that DTI is pushing the APR high, because on the same \
DTI, for example 0.01, we can see APR ranging from 5% to 40% and the same on \
almost every DTI level.

DTI interquartile range
```{r echo=FALSE}
summary(prosper_loans$DebtToIncomeRatio)
```

```{r echo=FALSE, warning=FALSE, message = FALSE}
ggplot(aes(x = BankcardUtilization, y = BorrowerAPR), data = prosper_loans)+
  geom_jitter(alpha = 2/5, shape = 21, fill = I("#FFA500"))+
  scale_x_log10()
```

From this plot we see little evidence that the utilization of the bank card \
have any effect on the borrower APR because as we can see that the APR ranges \ 
from 5% to 40% on every single bank card utilization ratio, but we can also \
notice that there is almost little borrowes above the 

```{r echo=FALSE, warning=FALSE}
ggplot(aes(x = MonthlyLoanPayment, y = BorrowerAPR), data = prosper_loans)+
  geom_point(alpha = 1/100)+
  scale_x_log10()
```

```{r}
summary(prosper_loans$MonthlyLoanPayment)
```

With 90% of loan payment between $130 and $370 per month, we can see that there \
is something going on in this plot, we dont see straight lines as we did in \
other variables and we can see on higher monhtly payments the APR doesnt go \
beyond 25% although we cant confirm because data there is little.


```{r echo=FALSE, warning=FALSE}
ggplot(aes(x = AmountDelinquent, y = BorrowerAPR), data = prosper_loans)+
  geom_jitter(alpha = 2/5, shape = 21, fill = I("#FFA500"))+
  scale_x_log10()
```

In the amount delinquent for borrowers we don't see a relationship with the \ 
borrower APR.

```{r echo=FALSE, warning=FALSE}
ggplot(aes(x = BorrowerState, y = BorrowerAPR), data = prosper_loans)+
  geom_boxplot()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

We can see some states median dropping below 20% and some above 20% but in \
general we dont see extreme APR rates in any of the states and those that might \
look low are mainly due to the smal number of loans so its difficult to reach \
conclusions.

```{r echo=FALSE}
ggplot(aes(x = Term, y = LoanOriginalAmount), data = prosper_loans)+
  geom_boxplot()
```

Looking at the 3 terms of loans provided, we can see a relationship between \
the size of the loan and the term of the loan which makes sense because \
usually bigger loans need more time to be paid which explains the difference \
in median which is $5,000 for the 36 month term and more than double $11,000.

```{r echo=FALSE}
ggplot(aes(x = BorrowersGroups, y = LoanOriginalAmount), data = prosper_loans)+
  geom_boxplot()
```

When comparing the delinquent borrowers to those whom are non-delinquent, we \ 
can see that delinquent borrowers median loan amount is $5,000 compared to\
around $7,000 for the the non-delinquent. We can also notice that the \
delinquent borrowers amount borrowered are less spread than non-delinquent \ 
with shorter tail and even the outliers are much conservative than the\
non-delinquent group.

```{r echo=FALSE, warning=FALSE}
ggplot(aes(y = BorrowerAPR, x = IncomeRange), data = subset(prosper_loans, IncomeRange != "NA" & IncomeRange != "$0"))+
  geom_boxplot(aes(color= IncomeRange))+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

After removing both the NA salaries and the zero salaries, \
we can see that the high earners have the lowers median APR of 18% and even \
the range of APR max is lower than that of the not-employed and lower salaries.\
We can see that the median of APR is dropping as salary increases assuring \ 
that a person's financial security is key to him or her getting a lower \
or higher APR.

```{r}
cor.test(prosper_loans$BorrowerRate, prosper_loans$BorrowerAPR, method = c("pearson"))
```

We can see the strong correlation between the 2 identical variables Borrower \
Rate and APR

```{r}
cor.test(prosper_loans$ProsperScore, prosper_loans$BorrowerAPR, method = c("pearson"))
```

We the negative correlation between a Borrower Prosper rating and the APR \
he or she gets because as the rating goes higher the APR decreases for low \
risk customers


-------------------------------------------------------------------------------

# Bivariate Analysis

-------------------------------------------------------------------------------

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

We can't see a clear relationship between most variables and the APR rate,\
but we know that Prosper tends to increase the borrowes APR if they think they \ 
are risky borrowers because when grouping delinquent and non-delinquent \
borrowers we noticed that there is 5% more APR on average between the 2 groups.\
This is clear when looking at the Prosper Rating given to the borrower day 1 \
and his/her APR. \
Also, we can see a relationship between the borrower APR and Income range, \
although couldn't see it with monthly salary but in the income range its clear \ 
that the higher the salary the lower the APR. \
Most important when it comes to the risky borrowers get a higher APR is in the \
APR distribution over years, because when the economy was bad, we saw high \
APR levels that went back to normal during the years after 2009.
Also, we can see that Interest rate has 99% correlation but only because these \
are almost the same variables so we can't consider it a correlation to be \
studied.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

We can see a relationship between the size of the loan and the loan term, \
because usually bigger loans tends to need more time to be paid but more \
interesting is that Prosper not only charged the delinquent borrowers with \
higher APR, they even gave them loans of smaller amounts, because as I noted \ 
in the above plot we can see that delinquent borrowers median loan amount is \
5,000 compared to around 7,000 for the the non-delinquent. \
Also we can see a rather strong negative correlation a borrower APR and the \
Prosper rating which proves that APR is used by Prosper to protect themselves \
against risky customers.

### What was the strongest relationship you found?

The strongest was between the Borrower APR and the Intereste Rate, but also \
we can see a relationship between the borrower status and the APR and also \
between his/her status and the size of the loan they will borrow and \
definitely the APR levels go down as salary increases making the borrower seem \
more secure.

-------------------------------------------------------------------------------

# Multivariate Plots Section

-------------------------------------------------------------------------------

```{r echo=FALSE, warning=FALSE}
ggplot(aes(y = BorrowerAPR,  x = ListingCreationYear), data = subset(prosper_loans, IncomeRange != "NA" & IncomeRange != "$0"))+
  geom_boxplot(aes(fill= IncomeRange))
```


```{r echo=FALSE, warning=FALSE}
table(prosper_loans$IncomeRange)
```

There are lots of things we can see in this plot, we can see that from 2013, \
prosper stopped accepting not-employed borrowers whom APR was the highest among
all groups. We can see that after 2011, APR levels for all groups started 
getting lower to reach its lowest levels in 2014.

```{r echo=FALSE, warning=FALSE}
ggplot(aes(x = BorrowerAPR), data = subset(prosper_loans, ProsperRating..Alpha. != "NA"))+
  geom_density(aes(color= ProsperRating..Alpha.))
```

This is amazing, we can see very clear that Prosper penalizes risky borrowers \
with high levels of APR, all over the plot HR (High Risk) is never visible \
until we reach 35%, and I believe these are the not-employed introduced in \
2013.\
lets zoom in on the levels to see how they are distributed \

```{r echo=FALSE, warning=FALSE}
ggplot(aes(x = BorrowerAPR), data = subset(prosper_loans, ProsperRating..Alpha. != "NA"))+
  geom_density(aes(color= ProsperRating..Alpha.))+
  ylim(0, 50)
```

After zooming in, we can see that almost some levels don't overlap, the "AA" \
sits on the left side with most of the borrowers getting less than 10%, \
then the "A" between 10% and 15% then the B between 15% and 20%.


```{r echo=FALSE, warning=FALSE}
ggplot(aes(x = BorrowerRate, y = BorrowerAPR), data = prosper_loans)+
  geom_boxplot()+
  facet_wrap(~BorrowersGroups)
```

The difference is clear between the borrower APR and Interest rate between \
delinquent borrowers and non-delinquent, so if Prosper think you are riskier \
you will get a higher APR.

```{r echo=FALSE, warning=FALSE}
ggplot(aes(y = BorrowerAPR, x = BorrowerRate, fill= ProsperRating..Alpha.), data = subset(prosper_loans, ProsperRating..Alpha. != "NA"))+
  stat_ellipse(geom = "polygon")
```

When plotting using confidence ellipses we can see the seperation between \
the different groups of borrowers rating, its clear that the low rik "AA" \
borrowers almost dominate the lowest Borrower Rate/APR an then the graph \
goes through different groups accordingly, we can see that the more a borrower \
is flagged as risky, the more we can see the groups intersect. We can also \
notice that the highest risk borrower "HR" is a very small group almost not \
visible on the plot.
 
```{r echo=FALSE, warning=FALSE}
ggplot(aes(x = BorrowerAPR), data = subset(prosper_loans, ProsperRating..Alpha. != "NA"))+
  geom_density(aes(fill= Term), alpha = 1/3)+
  facet_wrap(~BorrowersGroups)
``` 

We can see lots of interesting things here, Its now clear that Prosper \ 
borrowers whom ended up to be delinquent where in the first place risky as \
they have the majority of the skyrocket APR levels. We can also see that the \ 
60 month term borrowers had a lower APR than the popular 36 month term \
borrowers in both groups.

```{r echo=FALSE, warning=FALSE}
ggplot(aes(x = BorrowerAPR), data = subset(prosper_loans, ListingCreationYear > 2010 & ListingCreationYear < 2015))+
  geom_density(aes(color=ListingCreationYear))
```

We know that the bulk of the borrowers are from years 2011 till 2013, for that \
the above years are only visible in the plot. \
Looking at the years after the financial crisis we can see the difference \
in the distribution of the Borrower APR expecially in 2013 when it started \
to become normal where almost half of the loans are above 20% and half \
almost below unlike 2011 where we see spikes in the density at 30% and 40% \ 
APR and same for 2012 but less moderate.\
I was reluctant to add 2014 because the data is not complete but if we look at \ 
what data we have now, it shows that APR dropped and is becoming skewed to the \
right where the majority of loans are in the 15% APR range which is either \
due to Prosper attracting people with better ratings or the general economical \
situation is recovering.


```{r echo=FALSE, warning=FALSE}
ggplot(aes(x = ProsperRating..Alpha., y = BorrowerRate), data = subset(prosper_loans, ProsperRating..Alpha. != "NA"))+
  geom_boxplot()+
  facet_wrap(~ListingCreationYear)
```

We can see how much APR levels has dropped if we look, we can \
see that the are no more outliers in 2014 and we can see that even the HR \
rated borrowers dont exceed at max the 30% APR, we can also notice that the \ 
IQR was big on every level but now in 2014 it has shrinked \
 
-------------------------------------------------------------------------------

# Multivariate Analysis

-------------------------------------------------------------------------------

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

It is clear more that a person income range affects the APR he or she get at \
Prosper. We can also see that the borrower rating affects the APR which \
gets us to conclude that the difference between the APR and interest is mainly \
due to the borrower riskiness.

### Were there any interesting or surprising interactions between features?

The interesting part was the APR levels dropped over the years which is \
something quite odd especially with financial institutes whom tend to usually \
add fees and interest rates as they grow bigger.

The surprising was the introduction of the not-employed people in 2013 with \
skyrocket APR.

------

-------------------------------------------------------------------------------

# Final Plots and Summary

-------------------------------------------------------------------------------

### Plot One
```{r echo=FALSE, warning=FALSE}
ggplot(aes(y = BorrowerAPR, x = IncomeRange), data = subset(prosper_loans, IncomeRange != "NA" & IncomeRange != "$0"))+
  geom_boxplot(aes(color= IncomeRange))+
  labs(x = "Income Range [$]", y = "Borrower APR [%]")+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

### Description One

I chose this to be the first because it might have explained maybe most of the \
variation in APR that is not explained by the interest rate. This here I \
believe affects the initial prosper rating. It is clear when the APR levels \ 
starts to go very high when the borrower monthly salary drops which can be \
double between a high earner and someone denoted as not-employed which allows \ 
us to confirm that the financial security of a borrowers highly influences \
his or her APR level.


### Plot Two
```{r echo=FALSE, warning=FALSE}
ggplot(aes(x = BorrowerAPR), data = subset(prosper_loans, ProsperRating..Alpha. != "NA"))+
  geom_density(aes(color= ProsperRating..Alpha.))+
  ylim(0, 50)+
  labs(x = "Borrower APR [%]")
```

### Description Two

This plot was very supporting because as we can see that the prosper rating \
given day 1 to the borrower affects the APR level he or she get, we can almost \
see exclusivity for the borrowers rating levels as if they are separate bin. \
For example the AA ratings never go beyond 10% in APR while the C rating sits \
between the 15% and the 20% as for the HR, we now know that these are \
"not-employed" borrowers from 2013 whom where mostly given APR levels of 35% \
as they are risky and might become delinquent.


### Plot Three
```{r echo=FALSE, warning=FALSE}
ggplot(aes(x = BorrowerAPR), data = subset(prosper_loans, ListingCreationYear > 2010 & ListingCreationYear < 2015))+
  geom_density(aes(color=ListingCreationYear))+
  labs(x = "Borrower APR [%]")
```

### Description Three

I wanted to concentrate on this because it was surprising that the APR level \
dropped and is still dropping at Prosper, I believe this can be due to many \
reasons but my theories are that probably Prosper is attracting less riskier \
highly paid borrowers (although they gave loans to non-employed borrowers in \
2013) and  due to the recovery of the US economy after the housing crisis in \
2008-2009.


------

# Reflection

I was interested in this dataset because I was interested in Prosper as a \ 
company but when looking at the number of variables and since there are 81, \ 
I was reluctant at first because there are too many to be analyzed and even \
choosing few variables wasn't easy, I had to repeat the project 3 times before \ 
finding a problem that was unique, interesting and which I can choose adequate \
number of variables but not too many.

Also the plotting wasn't easy as the size of the data is big, and mostly was \ 
skewed which caused plots to be useless, even I was not satisfied with with \
ggpair plot because the variables that I chose was too much to be plotted at \
once.

Maybe I should have studied the riskiness of the borrower rather than the \
APR difference. It started with me because I thought APR was the same for all \
borrowers, but when I came across the difference in APR I wanted to know why, \
to be honest I was dreaming of this highly correlated continuous variables that \
will popup from somewhere and just explain it all but it seems in real life EDA \ 
isn't straight forward. 

I believe I have concluded that the borrower salary has \
a huge affect on the APR he or she gets but I don't think this alone helps \
because as we can see APR levels dropped over the years.
I believe Prosper has system that measures the borrower riskiness because as \
we saw over and over again, borrowers who are delinquent tended to have higher \ 
APR levels.

Now when I look at the variables and the prosper dataset I feel very familiar \
with it and I now have an idea how loans work, I know what affects the amount \
or the APR levels, also I know that its not only for low income borrowers, I \
also know that it has dramatically changed over time and that the average \
borrowerd amount is around $7,000.

Maybe if we add how salaries changed over the years, this could have been quite \
an idea, because there we can know if Prosper is attracting more secure \
borrowers or is it the economy, also if we can add a benchmark for what APR \
levels other P2P companies are providing and even what bank are providing in \
terms of micro loans. 

I think there are lots of issues that can be investigated and even after \
I feel I left lots of issues that could have been investigated especially how \ 
Prosper rates its borrowers upon approval of the loan which came late to me \
when I was almost done with the APR analysis and couldn't go back to start \
again for now.





